---
title: "Soy Deforestation: Attribution Methods Analysis"
author: "florian gollnow"
date: "May 20, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    keep-md: true
execute:
  message: false
  warning: false
---

# Problem definition

Trase subnational data reports commodity deforestation as the summed deforestation over a specific time period that occurred on the land used for commodity production. This is different from DeDuCE, classifying assigning each deforestation event to a specific commodity. 

DeDuCE approach is partly considered superior, because it is true to the annually observed deforestation (no double counting), allowing single and multi year reporting of commodity deforestation (multi year is not supported by Trase CDO sums, because one deforestation event would be counted multiple times.)

DeDuCE reports deforestation at the deforestation event, which may be unsuitable for Trase CDO, aiming to report the deforestation embedded in the harvest. 
However, DeDuCE metrics can easily be shifted to the harvest year instead of deforestation year, enabling reporting embedded deforestation.

## Suggested solution detailed below
Add additional metric following DeDuCE deforestation classification approach and report commodity deforestation
- At deforestation event year (note, recent years will undercount deforestation)
- At harvest year (note, by moving to harvest year, we can report the most recent year)


# Exploring the new deforestation metrics
This document compares Trase commodity deforestation attribution methods, with DeDuCE forward looking approach, and compares results to original DeDuCE results. 

Aim is to suggest a more comparable metric for Trase to adopt in its context impacts reporting, comparable to DeDuCE and annual deforestation reporting. 

## Setup and data ingestion

In this section, we load the required libraries for spatial analysis, data manipulation, and visualization, and import our annual soy metrics datasets.

```{r}
#| label: setup
library(tidyverse)
library(slider)
library(arrow)
library(geofacet)
library(scales)
library(sf)
library(zoo)

# Load datasets
soy_br <- read_parquet("~/documents/data/annual_metrics/soy_annual_br.parquet")
soy_states <- read_parquet(
  "~/documents/data/annual_metrics/soy_annual_br_states.parquet"
)
soy_supply_chain_2022_v2 <- read_parquet(
  "~/documents/data/annual_metrics/soy_2022_post_embedding_quants_v2.parquet"
)
```

# Country-level analysis
I begin by evaluating how different methodological choices impact estimated soy-driven deforestation at the national level. Specifically, I compare:

- 5-year Sum: Cumulative backward-looking deforestation.
- 5-year Annualized: Backward-looking annualized deforestation.
- 5-year / 3-year Harvest Attribution: Forward-looking spatial attribution windows.

## Annual attribution comparison
```{r}
#| label: country-annual-comparison
#| fig-width: 9
#| fig-height: 6
ggplot(
  soy_br |>
    filter(year >= 2010 & year <= 2024) |>
    filter(
      variable %in%
        c(
          "soy_def_5y_back",
          "soy_def_5y_annualized_back",
          "soy_def_harvest5y",
          "soy_def_harvest3y"
        )
    ) |>
    mutate(
      variable = case_when(
        variable == "soy_def_5y_back" ~ "Trase-5y-sum",
        variable == "soy_def_5y_annualized_back" ~ "Trase-5y-annualized",
        variable == "soy_def_harvest5y" ~ "Soy-def-@harvest_5y",
        variable == "soy_def_harvest3y" ~ "Soy-def-@harvest_3y",
        .default = variable
      )
    ),
  aes(year, ha / 10000, color = fct_reorder(variable, ha, .desc = TRUE))
) +
  geom_line(lwd = 1.5) +
  labs(
    title = "Soy Deforestation: Comparing Attribution Methods (Annual)",
    y = "Deforestation (kha)",
    x = "Year",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Key insights:

- 5-year Sum inherently double-counts annual deforestation over the evaluation window.
- 5-year Annualized undercounts actual annual deforestation. This occurs because the conversion from cleared land to soy commonly does not happen in the immediate first year after deforestation; dividing uniformly by 5 dilutes the calculated impact.
- 3-year vs. 5-year forward attribution windows yields highly similar trends, though the 5-year window remains expectedly higher.

## Amortized 5-year estimation
To smooth out annual volatility, I calculate a 5-year right-aligned rolling mean (amortization) for the forward-looking harvest attribution variables.

```{r}
#| label: country-amortization
soy_br_5ymean <- soy_br |>
  filter(variable %in% c("soy_def_harvest5y", "soy_def_harvest3y")) |>
  ungroup() |>
  group_by(variable) |>
  arrange(year, .by_group = TRUE) |>
  mutate(amortized_5 = rollmean(ha, k = 5, align = "right", fill = NA)) |>
  transmute(
    year = year,
    variable = paste0(variable, "_5y_amort"),
    ha = amortized_5
  )

soy_br_amort <- soy_br |> bind_rows(soy_br_5ymean)
```
```{r}
#| label: country-amortized-plot
#| fig-width: 9
#| fig-height: 6
ggplot(
  soy_br_amort |>
    filter(year >= 2013 & year <= 2024) |>
    filter(
      variable %in%
        c(
          "soy_def_5y_back",
          "soy_def_5y_annualized_back",
          "soy_def_harvest5y_5y_amort",
          "soy_def_harvest3y_5y_amort"
        )
    ) |>
    mutate(
      variable = case_when(
        variable == "soy_def_5y_back" ~ "Trase-5y-sum",
        variable == "soy_def_5y_annualized_back" ~ "Trase-5y-annualized",
        variable == "soy_def_harvest5y_5y_amort" ~ "Soy-def-@harvest_5y_amort",
        variable == "soy_def_harvest3y_5y_amort" ~ "Soy-def-@harvest_3y_amort",
        .default = variable
      )
    ),
  aes(year, ha / 10000, color = fct_reorder(variable, ha, .desc = TRUE))
) +
  geom_line(lwd = 1.5) +
  labs(
    title = "Soy Deforestation: Comparing Attribution Methods (5y Amortized)",
    y = "Deforestation (kha)",
    x = "Year",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 2))
```

## Key insights:

- amortization smoothes the reported deforestation estimates. Decision if annual or amortized metrics are better to show needed.
- from data perspective, amortization might provide more consistent and robust reporting
  - does not pick up changing trends as quick, but also more robust to data issues
  - from a user understanding a more simple annual metric.


relevant context information:
- Deduce dashboard reports annual deforestation
- Trase factsheets use 5 year amortized DeDuCE deforestation

## Percentage differences Between Methods
To quantify the divergence between backward-looking and forward-looking metrics, I also look at the relative percentage differences.
```{r}
#| label: country-percentage-diff
#| fig-width: 10
#| fig-height: 5
soy_br_amort_wide <- soy_br_amort |>
  pivot_wider(id_cols = year, names_from = variable, values_from = ha)

soy_br_amort_perc <- soy_br_amort_wide |>
  filter(year >= 2013) |>
  transmute(
    year = year,
    diff_perc_5y_Trase_ann_5y_ff_amort = ((soy_def_5y_annualized_back -
      soy_def_harvest5y_5y_amort) /
      soy_def_harvest5y_5y_amort) *
      100,
    diff_perc_5y_Trase_ann_3y_ff_amort = ((soy_def_5y_annualized_back -
      soy_def_harvest3y_5y_amort) /
      soy_def_harvest3y_5y_amort) *
      100,
    diff_perc_5y_ff_3y_ff_amort = ((soy_def_harvest5y_5y_amort -
      soy_def_harvest3y_5y_amort) /
      soy_def_harvest3y_5y_amort) *
      100
  )

soy_br_amort_perc_l <- soy_br_amort_perc |>
  pivot_longer(cols = -year, names_to = "variable", values_to = "percent")

ggplot(soy_br_amort_perc_l, aes(year, percent)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  facet_wrap(~ fct_reorder(variable, percent)) +
  labs(
    title = "Relative Percentage Difference Between Attribution Methods",
    x = "Year",
    y = "Difference (%)"
  ) +
  scale_x_continuous(breaks = breaks_pretty()) +
  theme_minimal()
```

## Key insights:

- Annualization of the Trase 5-year sum would miss between 20% and 80% of soy deforestation annually, depending heavily on the reference year and whether a 3- or 5-year attribution window is chosen.
- A 5-year forward-looking window maps roughly 22% to 30% more soy deforestation than a 3-year forward-looking window.

# Forward looking attribution vs deduce
As the intention is to align methods between Trase subnational context and DeDuCE, it is important to understand the difference between attribution to deforestation vs harvest year

```{r}
#| label: country-annual-comparison 1
#| fig-width: 9
#| fig-height: 6
ggplot(
  soy_br |>
    filter(year >= 2010 & year <= 2024) |>
    filter(
      variable %in%
        c(
          "soy_def_5y_back",
          #"soy_def_harvest5y",
          "soy_def_harvest3y",
          "soy_def_def3y",
          "deduce_unarmotized",
          "soy_def_def3y_gfc"
        )
    ) |>
    mutate(
      variable = case_when(
        variable == "soy_def_5y_back" ~ "Trase-5y-sum",
        #variable == "soy_def_harvest5y" ~ "Soy-def-@harvest_5y",
        variable == "soy_def_harvest3y" ~ "Soy-def-@harvest_3y",
        variable == "soy_def_def3y" ~ "Soy-def-@def_3y",
        variable == "deduce_unarmotized" ~ "DeDuCE_unarmotized",
        variable == "soy_def_def3y_gfc" ~ "DeDuCE_spatial",
        .default = variable
      )
    ),
  aes(year, ha / 10000, color = fct_reorder(variable, ha, .desc = TRUE))
) +
  geom_line(lwd = 1.5) +
  labs(
    title = "Soy Deforestation: Comparing Trase and DeDuCE",
    y = "Deforestation (kha)",
    x = "Year",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
ggplot(
  soy_br |>
    filter(year >= 2010 & year <= 2024) |>
    filter(
      variable %in%
        c(
          #"soy_def_5y_back",
          #"soy_def_harvest5y",
          #"soy_def_harvest3y",
          "soy_def_def3y",
          "deduce_unarmotized",
          "soy_def_def3y_gfc"
        )
    ) |>
    mutate(
      variable = case_when(
        # variable == "soy_def_5y_back" ~ "Trase-5y-sum",
        #variable == "soy_def_harvest5y" ~ "Soy-def-@harvest_5y",
        #variable == "soy_def_harvest3y" ~ "Soy-def-@harvest_3y",
        variable == "soy_def_def3y" ~ "Soy-def-@def_3y",
        variable == "deduce_unarmotized" ~ "DeDuCE_unarmotized",
        variable == "soy_def_def3y_gfc" ~ "DeDuCE_spatial",
        .default = variable
      )
    ),
  aes(year, ha / 10000, color = fct_reorder(variable, ha, .desc = TRUE))
) +
  geom_line(lwd = 1.5) +
  labs(
    title = "Soy Deforestation: Comparing Trase and DeDuCE at deforestation event",
    y = "Deforestation (kha)",
    x = "Year",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## key difference
- difference of trends due to different attribution year DeDuCE attributes to deforestation, while Trase commonly aims to attribute to harvest. 
- overall difference in total amount are due to differences in forest definition and data deforestation data used


# State-Level Analysis
I now scale the calculations down to individual Brazilian states to determine if sub-national trends and patterns align with national averages.
```{r}
#| label: state-amortization
soy_states_5ymean <- soy_states |>
  filter(variable %in% c("soy_def_harvest5y", "soy_def_harvest3y")) |>
  ungroup() |>
  group_by(variable, code, name, abbreviation) |>
  arrange(year, .by_group = TRUE) |>
  mutate(amortized_5 = rollmean(ha, k = 5, align = "right", fill = NA)) |>
  transmute(
    year = year,
    code = code,
    name = name,
    abbreviation = abbreviation,
    variable = paste0(variable, "_5y_amort"),
    ha = amortized_5
  )

soy_states_amort <- soy_states |> bind_rows(soy_states_5ymean)
```

## State-level: Amortized
```{r}
#| label: state-facet-geo
#| fig-width: 11
#| fig-height: 8
ggplot(
  soy_states_amort |>
    filter(year >= 2013) |>
    filter(
      variable %in%
        c(
          "soy_def_5y_back",
          "soy_def_5y_annualized_back",
          "soy_def_harvest5y_5y_amort",
          "soy_def_harvest3y_5y_amort"
        )
    ) |>
    mutate(
      variable = case_when(
        variable == "soy_def_5y_back" ~ "Trase-5y-sum",
        variable == "soy_def_5y_annualized_back" ~ "Trase-5y-annualized",
        variable == "soy_def_harvest5y_5y_amort" ~ "Soy-def-@harvest_5y_amort",
        variable == "soy_def_harvest3y_5y_amort" ~ "Soy-def-@harvest_3y_amort",
        .default = variable
      )
    ),
  aes(year - 2000, ha / 10000, color = fct_reorder(variable, ha, .desc = TRUE))
) +
  geom_line() +
  scale_x_continuous(breaks = breaks_pretty()) +
  labs(x = "Years (Since 2000)", y = "Deforestation (kha)", color = NULL) +
  facet_geo(~abbreviation, grid = "br_states_grid1") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
## State-level: amortized vs non-amortized
```{r}
#| fig-width: 10
#| fig-height: 7
ggplot(
  soy_states_amort |>
    filter(year >= 2013) |>
    filter(
      variable %in%
        c(
          "soy_def_5y_annualized_back",
          "soy_def_harvest5y_5y_amort",
          "soy_def_harvest3y_5y_amort"
        )
    ) |>
    mutate(
      variable = case_when(
        variable == "soy_def_5y_annualized_back" ~ "Trase-5y-annualized",
        variable == "soy_def_harvest5y_5y_amort" ~ "Soy-def-@harvest_5y_amort",
        variable == "soy_def_harvest3y_5y_amort" ~ "Soy-def-@harvest_3y_amort",
        .default = variable
      )
    ),
  aes(year - 2000, ha / 10000, color = fct_reorder(variable, ha, .desc = TRUE))
) +
  geom_line() +
  scale_x_continuous(breaks = breaks_pretty()) +
  labs(x = "Years (Since 2000)", y = "Deforestation (kha)", color = NULL) +
  facet_geo(~abbreviation, grid = "br_states_grid1") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Non-Amortized (All Metrics)
```{r}
#| fig-width: 10
#| fig-height: 7
ggplot(
  soy_states_amort |>
    filter(year >= 2013) |>
    filter(
      variable %in%
        c(
          "soy_def_5y_back",
          "soy_def_5y_annualized_back",
          "soy_def_harvest5y",
          "soy_def_harvest3y"
        )
    ) |>
    mutate(
      variable = case_when(
        variable == "soy_def_5y_back" ~ "Trase-5y-sum",
        variable == "soy_def_5y_annualized_back" ~ "Trase-5y-annualized",
        variable == "soy_def_harvest5y" ~ "Soy-def-@harvest_5y",
        variable == "soy_def_harvest3y" ~ "Soy-def-@harvest_3y",
        .default = variable
      )
    ),
  aes(year - 2000, ha / 10000, color = fct_reorder(variable, ha, .desc = TRUE))
) +
  geom_line() +
  scale_x_continuous(breaks = breaks_pretty()) +
  labs(x = "Years (Since 2000)", y = "Deforestation (kha)", color = NULL) +
  facet_geo(~abbreviation, grid = "br_states_grid1") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Non-Amortized (Excl. 5y-Sum)
```{r}
#| fig-width: 10
#| fig-height: 7
ggplot(
  soy_states_amort |>
    filter(year >= 2013) |>
    filter(
      variable %in%
        c(
          "soy_def_5y_annualized_back",
          "soy_def_harvest5y",
          "soy_def_harvest3y"
        )
    ) |>
    mutate(
      variable = case_when(
        variable == "soy_def_5y_annualized_back" ~ "Trase-5y-annualized",
        variable == "soy_def_harvest5y" ~ "Soy-def-@harvest_5y",
        variable == "soy_def_harvest3y" ~ "Soy-def-@harvest_3y",
        .default = variable
      )
    ),
  aes(year - 2000, ha / 10000, color = fct_reorder(variable, ha, .desc = TRUE))
) +
  geom_line() +
  scale_x_continuous(breaks = breaks_pretty()) +
  labs(x = "Years (Since 2000)", y = "Deforestation (kha)", color = NULL) +
  facet_geo(~abbreviation, grid = "br_states_grid1") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# key insight:
- trends and patterns are consistent across attribution methods


# Supply Chain & Trader Exposure (2022)
In this final section, we look at supply chain allocations for the year 2022 to isolate the top 10 exporters, production regions, and importing country groups exposing themselves to deforestation risks under alternative accounting frameworks.
```{r}
#| label: supply-chain-aggregation
# Exporter Aggregation
soy_supply_chain_2022_v2_exp <- soy_supply_chain_2022_v2 |>
  group_by(EXPORTER_GROUP) |>
  summarize(across(ends_with("_exp"), ~ sum(.x, na.rm = TRUE)))

# Production Region Aggregation
soy_supply_chain_2022_v2_reg <- soy_supply_chain_2022_v2 |>
  group_by(LVL6_NAME_PROD) |>
  summarize(across(ends_with("_exp"), ~ sum(.x, na.rm = TRUE)))

# Importer Country Aggregation
soy_supply_chain_2022_v2_imp <- soy_supply_chain_2022_v2 |>
  group_by(COUNTRY_GROUP) |>
  summarize(across(ends_with("_exp"), ~ sum(.x, na.rm = TRUE)))
```
## Top 10 Trader Exposure Lists
Below we compare how different metrics rearrange or weigh the risk profiles of top exporters.
```{r}
cat("Trase 5-Year Sum Deforestation\n")
soy_supply_chain_2022_v2_exp |>
  arrange(desc(soy_def_5y_back_exp)) |>
  select(EXPORTER_GROUP, soy_def_5y_back_exp) |>
  filter(!EXPORTER_GROUP %in% c("PROCESSED DOMESTICALLY", "UNKNOWN")) |>
  top_n(10)

cat("Trase Forward-Looking 5-Year Amortized\n")
soy_supply_chain_2022_v2_exp |>
  arrange(desc(soy_def_harvest5y_5y_amort_exp)) |>
  select(EXPORTER_GROUP, soy_def_harvest5y_5y_amort_exp) |>
  filter(!EXPORTER_GROUP %in% c("PROCESSED DOMESTICALLY", "UNKNOWN")) |>
  top_n(10)

cat("Trase Forward-Looking 3-Year Amortized\n")
soy_supply_chain_2022_v2_exp |>
  arrange(desc(soy_def_harvest3y_5y_amort_exp)) |>
  select(EXPORTER_GROUP, soy_def_harvest3y_5y_amort_exp) |>
  filter(!EXPORTER_GROUP %in% c("PROCESSED DOMESTICALLY", "UNKNOWN")) |>
  top_n(10)

cat("Trase Forward-Looking 3-Year\n")
soy_supply_chain_2022_v2_exp |>
  arrange(desc(soy_def_harvest3y_exp)) |>
  select(EXPORTER_GROUP, soy_def_harvest3y_exp) |>
  filter(!EXPORTER_GROUP %in% c("PROCESSED DOMESTICALLY", "UNKNOWN")) |>
  top_n(10)
```

### key insights:
- same 5 Traders with highest number of deforestation
- order chchanges for 2 companies only

Top 10 Regional Exposure Lists (Production Level 6)
```{r}
cat("Trase 5-Year Sum Deforestation\n")
soy_supply_chain_2022_v2_reg |>
  arrange(desc(soy_def_5y_back_exp)) |>
  select(LVL6_NAME_PROD, soy_def_5y_back_exp) |>
  filter(LVL6_NAME_PROD != "UNKNOWN") |>
  top_n(10)

cat("Trase Forward-Looking 5-Year Amortized\n")
soy_supply_chain_2022_v2_reg |>
  arrange(desc(soy_def_harvest5y_5y_amort_exp)) |>
  select(LVL6_NAME_PROD, soy_def_harvest5y_5y_amort_exp) |>
  filter(LVL6_NAME_PROD != "UNKNOWN") |>
  top_n(10)

cat("Trase Forward-Looking 3-Year Amortized\n")
soy_supply_chain_2022_v2_reg |>
  arrange(desc(soy_def_harvest3y_5y_amort_exp)) |>
  select(LVL6_NAME_PROD, soy_def_harvest3y_5y_amort_exp) |>
  filter(LVL6_NAME_PROD != "UNKNOWN") |>
  top_n(10)


cat("Trase Forward-Looking 3-Year\n")
soy_supply_chain_2022_v2_reg |>
  arrange(desc(soy_def_harvest3y_exp)) |>
  select(LVL6_NAME_PROD, soy_def_harvest3y_exp) |>
  filter(LVL6_NAME_PROD != "UNKNOWN") |>
  top_n(10)
```

### key insights:
- same 5 Traders with highest number of deforestation
- order chchanges for 2 companies only


## Top 10 Importer Exposure Lists (Country Groups)

```{r}
cat("Trase 5-Year Sum Deforestation\n")
soy_supply_chain_2022_v2_imp |>
  arrange(desc(soy_def_5y_back_exp)) |>
  select(COUNTRY_GROUP, soy_def_5y_back_exp) |>
  filter(!COUNTRY_GROUP %in% c("BRAZIL", "UNKNOWN COUNTRY EUROPEAN UNION")) |>
  top_n(10)

cat("Trase Forward-Looking 5-Year Amortized\n")
soy_supply_chain_2022_v2_imp |>
  arrange(desc(soy_def_harvest5y_5y_amort_exp)) |>
  select(COUNTRY_GROUP, soy_def_harvest5y_5y_amort_exp) |>
  filter(!COUNTRY_GROUP %in% c("BRAZIL", "UNKNOWN COUNTRY EUROPEAN UNION")) |>
  top_n(10)

cat("Trase Forward-Looking 3-Year Amortized\n")
soy_supply_chain_2022_v2_imp |>
  arrange(desc(soy_def_harvest3y_5y_amort_exp)) |>
  select(COUNTRY_GROUP, soy_def_harvest3y_5y_amort_exp) |>
  filter(!COUNTRY_GROUP %in% c("BRAZIL", "UNKNOWN COUNTRY EUROPEAN UNION")) |>
  top_n(10)

cat("Trase Forward-Looking 3-Year\n")
soy_supply_chain_2022_v2_imp |>
  arrange(desc(soy_def_harvest3y_exp)) |>
  select(COUNTRY_GROUP, soy_def_harvest3y_exp) |>
  filter(!COUNTRY_GROUP %in% c("BRAZIL", "UNKNOWN COUNTRY EUROPEAN UNION")) |>
  top_n(10)
```

### Key insights:
- small changes in deforestation exposure change rankings slightly - but not dramatically.