---
title: "Soy Deforestation: Attribution Methods Analysis"
author: "florian gollnow"
date: "May 20, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    keep-md: true
execute:
  message: false
  warning: false
---

# Problem definition

Trase subnational data reports commodity deforestation as the summed deforestation over a specific time period that occurred on the land used for commodity production. This is different from DeDuCE, classifying assigning each deforestation event to a specific commodity. 

DeDuCE approach is partly considered superior, because it is true to the annually observed deforestation (no double counting), allowing single and multi year reporting of commodity deforestation (multi year is not supported by Trase CDO sums, because one deforestation event would be counted multiple times.)

DeDuCE reports deforestation at the deforestation event, which may be unsuitable for Trase CDO, aiming to report the deforestation embedded in the harvest. 
However, DeDuCE metrics can easily be shifted to the harvest year instead of deforestation year, enabling reporting embedded deforestation.

## Suggested solution detailed below
Add additional metric following DeDuCE deforestation classification approach and report commodity deforestation
- At deforestation event year (note, recent years will undercount deforestation)
- At harvest year (note, by moving to harvest year, we can report the most recent year)


# Exploring the new deforestation metrics
This document compares Trase commodity deforestation attribution methods, with DeDuCE forward looking approach, and compares results to original DeDuCE results. 

Aim is to suggest a more comparable metric for Trase to adopt in its context impacts reporting, comparable to DeDuCE and annual deforestation reporting. 

## Setup and data ingestion

In this section, we load the required libraries for spatial analysis, data manipulation, and visualization, and import our annual soy metrics datasets.

```{r}
#| label: setup
library(tidyverse)
library(slider)
library(arrow)
library(geofacet)
library(scales)
library(sf)
library(zoo)

# Load datasets
soy_br <- read_parquet(
  "~/documents/data/annual_metrics/soy_annual_br_v3.parquet"
)
soy_states <- read_parquet(
  "~/documents/data/annual_metrics/soy_annual_br_states_v3.parquet"
)
soy_supply_chain_2022_v2 <- read_parquet(
  "~/documents/data/annual_metrics/soy_2022_post_embedding_quants_v3.parquet"
)
```

# Country-level analysis
I begin by evaluating how different methodological choices impact estimated soy-driven deforestation at the national level. Specifically, I compare:

- 5-year Sum: Cumulative backward-looking deforestation.
- 5-year Annualized: Backward-looking annualized deforestation.
- 5-year / 3-year Harvest Attribution: Forward-looking spatial attribution windows.

## Annual attribution comparison
```{r}
#| label: country-annual-comparison
#| fig-width: 9
#| fig-height: 6
ggplot(
  soy_br |>
    filter(year >= 2014 & year <= 2024) |>
    filter(
      variable %in%
        c(
          "soy_def_5y_back",
          "soy_def_harvest3y_5y_amortized" #,
          #"deduce_unarmotized_5y_amortized"
        )
    ) |>
    mutate(
      variable = case_when(
        variable == "soy_def_5y_back" ~ "Trase-5y-sum",
        variable ==
          "soy_def_harvest3y_5y_amortized" ~ "Soy-def-@harvest_3y_amortized",
        variable == "deduce_unarmotized_5y_amortized" ~ "DeDuCE_amortized",
        .default = variable
      )
    ),
  aes(year, ha / 1000, color = fct_reorder(variable, ha, .desc = TRUE))
) +
  geom_line(lwd = 1.5) +
  ylim(c(0, 1100)) +
  labs(
    title = "Soy Deforestation: Comparing 5 year sum vs annual deforestation",
    y = "Deforestation (kha)",
    x = "Year",
    color = NULL
  ) +
  scale_x_continuous(breaks = breaks_pretty()) +

  theme(legend.position = "bottom")


ggplot(
  soy_br |>
    filter(year >= 2014 & year <= 2024) |>
    filter(
      variable %in%
        c(
          "soy_def_5y_back",
          "soy_def_harvest3y_5y_amortized",
          "deduce_unarmotized_5y_amortized"
        )
    ) |>
    mutate(
      variable = case_when(
        variable == "soy_def_5y_back" ~ "Trase-5y-sum",
        variable ==
          "soy_def_harvest3y_5y_amortized" ~ "Soy-def-@harvest_3y_amortized",
        variable == "deduce_unarmotized_5y_amortized" ~ "DeDuCE_amortized",
        .default = variable
      )
    ),
  aes(year, ha / 1000, color = fct_reorder(variable, ha, .desc = TRUE))
) +
  geom_line(lwd = 1.5) +
  ylim(c(0, 1100)) +
  labs(
    title = "Soy Deforestation: Comparing 5 year sum vs annual vs DeDuCE",
    y = "Deforestation (kha)",
    x = "Year",
    color = NULL
  ) +
  scale_x_continuous(breaks = breaks_pretty()) +

  theme(legend.position = "bottom")
```

## Key insights:

- 5-year Sum inherently double-counts  deforestation over the evaluation window.
- 3-year  forward attribution windows yields highly similar trends, though the 5-year window remains expectedly higher.
- DeDuCE is lower , due to different forest definition

## Percentage differences between methods
To quantify the divergence between backward-looking and forward-looking metrics, I also look at the relative percentage differences.
```{r}
#| label: country-percentage-diff
#| fig-width: 10
#| fig-height: 5
soy_br_amort_wide <- soy_br |>
  pivot_wider(id_cols = year, names_from = variable, values_from = ha)

soy_br_amort_perc <- soy_br_amort_wide |>
  filter(year >= 2013) |>
  transmute(
    year = year,
    diff_perc_ff_3y_ff_amort_5y_sum = ((soy_def_harvest3y_5y_amortized -
      soy_def_5y_back) /
      soy_def_5y_back) *
      100,
    diff_perc_DeDuCE_5y_amort_ff_3y_ff_amort_ = ((deduce_unarmotized_5y_amortized -
      soy_def_harvest3y_5y_amortized) /
      soy_def_harvest3y_5y_amortized) *
      100,
    diff_perc_DeDuCE_5y_amort_ff_def_3y_ff_amort_ = ((deduce_unarmotized_5y_amortized -
      soy_def_def3y_5y_amortized) /
      soy_def_def3y_5y_amortized) *
      100
  )

soy_br_amort_perc_l <- soy_br_amort_perc |>
  pivot_longer(cols = -year, names_to = "variable", values_to = "percent")

ggplot(
  soy_br_amort_perc_l |> filter(variable == "diff_perc_ff_3y_ff_amort_5y_sum"),
  aes(year, percent)
) +
  geom_bar(stat = "identity", fill = "steelblue") +
  facet_wrap(~ fct_reorder(variable, percent)) +
  labs(
    title = "Percentage difference between 5 year sum and annual soy deforestation",
    x = "Year",
    y = "Difference (%)"
  ) +
  scale_x_continuous(breaks = breaks_pretty()) +
  ylim(c(-100, 0))

ggplot(
  soy_br_amort_perc_l |>
    filter(variable == "diff_perc_DeDuCE_5y_amort_ff_3y_ff_amort_"),
  aes(year, percent)
) +
  geom_bar(stat = "identity", fill = "steelblue") +
  facet_wrap(~ fct_reorder(variable, percent)) +
  labs(
    title = "Percentage difference between DeDuCE sum and annual soy deforestation",
    x = "Year",
    y = "Difference (%)"
  ) +
  scale_x_continuous(breaks = breaks_pretty()) +
  ylim(c(-100, 0))


ggplot(
  soy_br_amort_perc_l |>
    filter(variable == "diff_perc_DeDuCE_5y_amort_ff_def_3y_ff_amort_"),
  aes(year, percent)
) +
  geom_bar(stat = "identity", fill = "steelblue") +
  facet_wrap(~ fct_reorder(variable, percent)) +
  labs(
    title = "Percentage difference between DeDuCE sum and annual soy deforestation @def",
    x = "Year",
    y = "Difference (%)"
  ) +
  scale_x_continuous(breaks = breaks_pretty()) +
  ylim(c(-100, 0))
```
